---
title: "Web graphics"
output:
  html_document:
    keep_md: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(base.url = "/SNR_R_Group/")
knitr::opts_chunk$set(fig.path = "figs/2016-10-04-WebGraphics/")
```

## Plots for the web

There has been increasing need and desirability to produce interactive graphics online.  News outlets, like the The Economist, New York Times, Vox, 538, Pew, and Quartz, routinely use interactive displays.

There are two packages (among several) that allow us to create interactive graphics.  The first is `ggivis`, which is based on `ggplot2` and the other is `googlevis`.  

ggvis follows the similar underlying theory of the grammar of graphics as ggplot2 but is expressed a little bit differently.  It incorporates aspects from `shiny` as well as `dplyr`.  

### `ggvis()`

```{r}
#install.packages("ggvis")

library(tidyverse)

library(ggvis)

```


Like `ggplot()`, every call to `ggvis` starts with the `ggvis()` statement.  

```{r}
plot_point<-mtcars %>%
            ggvis(x = ~wt, y = ~mpg) %>%
            layer_points()

view_static(plot_point)
```

