---
title: "Applied Multivarite:  Intro to Data Wrangling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This class covers an introduction to wrangling and summarizing data using the functions from `tidyr` and `dplyr`.  This is only meant to serve as a brief introduction, as we will keep developing these skills throughout the semester.  I have a similar lesson on data wrangling from a previous semester [2016-09-23-Wrangling](https://chrischizinski.github.io/SNR_R_Group/2016-09-23-Wrangling) and information on joins [2016-09-29-Joining_Data_Sets](https://chrischizinski.github.io/SNR_R_Group/2016-09-29-Joining_Data_Sets).

[R script from class](https://github.com/chrischizinski/SNR_R_Group/blob/master/R/2017-09-01-Rclass_dplyr.R)


[Answers to last week challenges](https://github.com/chrischizinski/SNR_R_Group/blob/master/R/2017-09-01-Challenges_week1.R)



## Weekly challenge

### The data

Work with the ecology data set from datacarpentry.  An explanation of the data set can be found [here](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html)

```{r message = FALSE, warning = FALSE}
library(tidyverse)

mydata <- read_csv("https://ndownloader.figshare.com/files/2292169")

glimpse(mydata)
```

## In-class challenges

*1. create a new column for Species (Genus + specific epithet)*

*2. Average weight of Male Rodents*

*3.put data into long form*


## TAKE home challenges

*1. How many animals caught in each plot type?*
*2. How many species are in each taxa?*
*3. Average weight of each species?*
*4. Average weight of each Sex of Each species?*