---
title: "Ecological Detective - Know your data"
output: html_document
---


```{r, echo = FALSE, message = FALSE}
# .libPaths("P:/RLibrary2")

library(tidyverse)

```

## Recreating Figure 2.1 in Ecological Detective 

There are many different hypotheses that can explain the basic relationship between two variables.  Figure 2.1 in the Ecologial Detective suggest 4 possible models.  The models have no parameter values.  Try to iteratively find the parameter values to get your figure to look like the one in Figure 2.1.

```{r}
S <- seq(1,15, by = 1)  # 1:15
Null_hypothesis<- 2.5

our_data <- data.frame(S = S, Null = Null_hypothesis)
our_data$Model_A <- 0.5 *S
our_data$Model_B <- (0.9 *S)/(1 + 0.1*S)
our_data$Model_C <- 1.8 *S * exp(-0.18*S)

plot_labels = data.frame(x = c(14,14,8, 4), y = c(7.5,5.5,2.75,3.5), labels = c("Model A", "Model B", "NULL", "Model C"))


ggplot() +
  geom_line(data = our_data,aes(x = S, y = Null), colour="black") +
  geom_line(data = our_data,aes(x = S, y = Model_A), colour="blue") +
  geom_line(data = our_data,aes(x = S, y = Model_B), colour="green") +
  geom_line(data = our_data,aes(x = S, y = Model_C), colour="red") +
  geom_text(data = plot_labels, aes(x=x, y= y, label = labels)) +
  coord_cartesian(ylim = c(0,9), xlim = c(0, 16), expand = FALSE) +
  scale_x_continuous(breaks = seq(0,16,by=2)) +
  scale_y_continuous(breaks = seq(1,8,by=1)) +
  labs(x = "Flock size", y = "Consumption rate") +
  theme_bw()
```


## Probability and probability models 

- Ecological data (and most other kinds of data) involve different levels of randomness
- Most models represent the mean of the population 
- Comparing models requires understanding the probability of individual observations (based on the distribution) 

- Most people understand the **normal** or Gaussian distribution, but htere are many, many different types of distributions  

**WORD OF ADVICE**:  Always plot your data 

```{r}
fish_data <- read_csv("https://raw.githubusercontent.com/chrischizinski/MWFWC_FishR/master/CourseMaterial/data/wrkshp_data.csv")

glimpse(fish_data)

fish_data %>% 
  select(WaterbodyCode:Age) %>% 
  mutate(Age = as.numeric(Age)) %>% 
  filter(!is.na(Age),
         WaterbodyCode == 4999,
         SpeciesCode %in% c(780)) -> FishAge  

glimpse(FishAge)

```


Let's look at distribution plots of `FishLength` and `Age`

```{r}
ggplot(data = FishAge) + 
  geom_histogram(aes(x = Age), binwidth = 1, fill = "green", colour = "black") +
  theme_bw()

ggplot(data = FishAge,aes(x = Age)) + 
  geom_histogram(aes(y = ..density..),binwidth = 1, fill = "green", colour = "black") +
  geom_density(colour = "red", fill = "red", alpha = 0.5) +
  theme_bw()

ggplot(data = FishAge) +
  geom_histogram(aes(x = FishLength), binwidth = 10, fill = "red", colour = "black" ) + 
  theme_bw()

ggplot(data = FishAge) +
  geom_violin(aes(x = "", y = FishLength), fill = "red", colour = "black" ) + 
  theme_bw()

ggplot(data = FishAge) +
  geom_point(aes(x = "", y = FishLength), position = "jitter") +
  geom_boxplot(aes(x = "", y = FishLength), fill = "red", colour = "black", alpha = 0.25, outlier.colour = NA) + 
  theme_bw()
  
```


