---
title: "Ecological Detective - Probability and probability models. Part 1"
output: html_document
---

```{r message = FALSE}
library(tidyverse)
```

## Classical Probability

Sources of the notes for this lecture are a combination of [Aho(2013)](http://www2.cose.isu.edu/~ahoken/book/) (Chapters 2 and 3) and Ecological Detective (Chapters 3 and 4).

- As we become familiar with the behavior of random variables, we may become aware of probabilistic patterns

### Disjoint

If two events can not occur simultaneously, then we call them **mutually exclusive** or **disoint**

```{r, echo = FALSE}
rect_1<-data.frame(outcome = factor(c(1,2)), xmin = c(0,3), xmax = c(3, 4), ymin = c(0,0), ymax = c(4,4))

ggplot(data = rect_1) + 
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = outcome), alpha = 0.25) +
  coord_equal(expand = FALSE) +
  geom_text(aes(x = c(1.5,3.6), y = c(2,2), label = c("P(A) = 0.75", "P(B) = 0.25"))) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.title = element_blank())
```

If for two outcomes *A* and *B*, we wanted to know the probability of the event *A* or *B* (expressed as:  \\(P( A \cup B) = P(A) + P(B) \\))

We can also think of events that are not mutually exclusive.  Suppose an organism has a probability of being in a habitat with environmental variable *A* with P = 0.3 and a habitat with environmental variable *B* with P = 0.3 and habitat with environmental variable *A* and *B* with P = 0.09

```{r, echo = FALSE}
rect_1<-data.frame(outcome = factor(c(1,2)), xmin = c(0,0.21), xmax = c(0.3, 0.51), ymin = c(0,0), ymax = c(0.51,0.51))

ggplot(data = rect_1) + 
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = outcome), alpha = 0.25) +
  coord_equal(expand = FALSE) +
  annotate("text", x = 0.1, y = 0.25, label = "P(A) = 0.3") +
  annotate("text", x = 0.4, y = 0.25, label = "P(B) = 0.3") +
  annotate("text", x = 0.25, y = 0.25, label = as.character(expression(paste("P(A",intersect("B"),") = 0.09"))), parse = TRUE) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text = element_blank(),
        axis.title = element_blank())
```

If *A* and *B* are not mutually exclusive we can sill calcualte the union of *A* and *B* as \\(P( A \cup B) = P(A) + P(B)  - P( A \cap B)\\))

### Independence 

- When event A occurs it does not affect the probability of B, then we say that A and B are independent.

## Conditional probability 

- There can be many events that are not independent 
- Suppose A is the outcome of a prey organism surviving a predator experiment on day 1 and B is the outcome of the same prey organism surviving the predator experiment the next day.   If the the prey can learn and alter it's behavior on day 1 then the outcome on day 2 is not independent.  We can denote this as \\( P(B\|A)\\) or the probability of *B* given *A*

## Odds

- Closely related to probability
- The ratio between the number of favorable outcomes to the number of unfavorable outcomes.  The odds of rolling a two on a dice are 1:5.  Total number of outcomes = 6, Number of favorable outcomes = 1, and unfavorable outcomes (6-1) = 5.

### Odds ratio and relative risk

- Ratio of the odds for two outcomes is their *odds ratio*
- Ratio of the probability of two events is *relative risk*

## Probability density functions

- A *probability distribution* assigns probabilities to outcomes from a random variable
- The mathematical expression \\(f(x)\\) that defines a probability distribution is called a *probability density function* or *pdf*
- The output of the *pdf* is called *density*

Example

```{r}
rand_vals<-data.frame(x = rnorm(10000))

ggplot(rand_vals) + 
  geom_histogram(aes(x = x), fill = "red", colour = "black", binwidth = 1) +
  theme_bw()

ggplot(rand_vals) + 
  geom_histogram(aes(x = x), fill = "red", colour = "black", binwidth = 0.25) +
  theme_bw()

ggplot(rand_vals) + 
  geom_histogram(aes(x = x), fill = "red", colour = "black", binwidth = 0.05) +
  theme_bw()

#  Add a density curve
ggplot(rand_vals,aes(x = x)) + 
  geom_histogram(aes(y = ..density..), fill = "red", colour = "black", binwidth = 0.05) +
  geom_density(size = 2, colour = "blue") +
  theme_bw()

```

1. Both discrete and continuous pdfs calculate a quantity called density. 
    - The “height” of any pdf “curve” at an outcome x will equal the density of x, given as \\( f(x) \\)
2. Density is equivalent to probability for a discrete pdf, but not for a continuous pdf. 


## Cumulative density functions

- *Cumulative distribution function* (*cdf)* for a random variable X is denoted \\(F(x)\\). 
    - cdf is obtained by summing (discrete) or integrating (continuous) the pdf between \\(-\inf\\) and an outcome *x*
    - cdf gives the lower tail probability \\( P(X \leq x) \\) for the both discrete and continuous random variables.

## Common distributions

### Discrete

#### Binomial

- The binomial distribution defines the probability of a particular number of successes given n independent and identically distributed Bernoulli trials
- two parameters: the number of trials (*n*) and the probability of success for a single trial ( \\( \pi \\))

Psuedo code 3.1

```{r}
p = 0.1
N = 10

P_N = (1-p)^N

psuedo_dat <- data.frame(k = 1:N)
psuedo_dat$P_N <- (factorial(N)/(factorial(psuedo_dat$k)*factorial((N - psuedo_dat$k)))) * (p^psuedo_dat$k)*(1-p)^(N -psuedo_dat$k) 
psuedo_dat$P_1 <- dbinom(x = 1:10, size = 10, p = 0.1)
psuedo_dat$P_2 <- dbinom(x = 1:10, size = 10, p = 0.2)
psuedo_dat$P_3 <- dbinom(x = 1:10, size = 10, p = 0.3)

psuedo_dat %>% 
  select(k, P_1:P_3) %>% 
  gather(probability, value, P_1:P_3) -> psuedo_dat.long

ggplot(psuedo_dat.long) + 
  geom_bar(aes(x = k, y = value, fill = probability, group = probability), stat = "identity", width = 0.75, position = "dodge") +
  coord_cartesian(ylim = c(0,0.40), xlim = c(0, 10), expand = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(y="density",x = "Number of fruit sampled") +
  theme_bw()

```

#### Poisson distribution 

- *Poisson distribution* is functionally similar to the binomial distribution
- the distribution gives the probability for a defined number of successes, x, given a series of independent trials. 
- As number of trials goes to infinity, the Poisson pdf will be equivalent to the binomial pdf
- No explicit maximum value for x because the Poisson distribution considers successes in the context of a  fixed success rate instead of a fixed sample size (binomial)
- One variable \\( \lambda \\) is the mean and variance, calculated as r (rate of succcesses) * t (unit of time)

Psuedo Code 3.2

Suppose the rate parameter *r* is 0.5 (probability a bird flys by in a given minute) and we want to watch over time *t* is 10. Given that we have and r of 0.5 and we will watch for t = 10, then we would expect that approximately 5 birds would fly by mu = r*t. 


```{r}
r = 0.5
t = 10
cutoff = 1 - 1e-9

sum_val <- exp(-r*t)

k = 1
pois_stor<-NULL

while(sum_val <= cutoff){
  
  p_kt<-((exp(-r * t) * (r*t)^k)/factorial(k))
  
  pois_stor<-rbind(pois_stor,data.frame(k = k, p_kt = p_kt))
  
  sum_val = sum_val +p_kt
  
  k<- k +1
  
}

# Final K
k
# Final p_kt
p_kt

pois_stor
```

Or we can use `dpois`

```{r}
pois_stor$p_kt1 <- dpois(1:23, r*t, log = FALSE)

ggplot(pois_stor) + 
  geom_bar(aes(x = k, y = p_kt1), stat = "identity", width = 0.75, fill = "lightblue") +
  coord_cartesian(ylim = c(0,0.20), xlim = c(0, 23), expand = FALSE) +
  scale_x_continuous(breaks = 1:23) +
  labs(y="density",x = "k") +
  theme_bw()
```

Calculate the cdf

```{r}
pois_stor$cdf_kt1 <-cumsum(pois_stor$p_kt1)
pois_stor
```

And plot the cdf

```{r}
ggplot(pois_stor) + 
  geom_bar(aes(x = k, y = cdf_kt1), stat = "identity", width = 0.75, fill = "lightblue") +
  coord_cartesian(ylim = c(0,1), xlim = c(0, 23), expand = FALSE) +
  scale_x_continuous(breaks = 1:23) +
  labs(y="cumulative density",x = "k") +
  theme_bw()
```

What is the probability of 4 birds flying by in 10 minutes?

```{r}
sum(dpois(0:4, r*t, log = FALSE))

ppois(4, r*t)
```

