---
title: "Ecological Detective - Relationships and probability"
output: html_document
---

## Exploring the relationship between two variables

First lets bring in the data from the previous lesson

```{r message=FALSE}
library(tidyverse)
library(broom)
```


```{r}
fish_data <- read_csv("https://raw.githubusercontent.com/chrischizinski/MWFWC_FishR/master/CourseMaterial/data/wrkshp_data.csv")

fish_data %>% 
  select(WaterbodyCode:Age) %>% 
  mutate(Age = as.numeric(Age)) %>% 
  filter(!is.na(Age),
         WaterbodyCode == 4999,
         SpeciesCode %in% c(780)) -> FishAge  

```


Now lets plot the basic relationship between age and length of this species. 

```{r}
ggplot(data = FishAge) +
  geom_point(aes(x = Age, y = FishLength), size = 3, alpha = 0.35, colour = "red") +
  theme_bw()
```

How does fish age relate to fish length?

1. Linear
2. Polynomial
3. Logarithmic
4. Non-linear

```{r}
# linear response
lm_mod <- lm(FishLength ~ Age, data = FishAge)
summary(lm_mod)

# polynomial response
poly_mod <- lm(FishLength ~ poly(Age,3), data = FishAge)
summary(poly_mod)

# logarithmic

log_mod <- lm(FishLength ~ log(Age +1), data = FishAge)
summary(log_mod)

# non linear

nl_mod <- nls(FishLength ~ exp(a + Age*b), start = list(a = 0, b = 1), data = FishAge)
summary(nl_mod)
```

Plot the curves to the data 

```{r}
newdata <- data.frame(Age = seq(0,8,by = 1))

lm_pred<- data.frame(model = "linear",augment(lm_mod, newdata = newdata))

poly_pred<- data.frame(model = "polynomial",augment(poly_mod, newdata = newdata))

log_pred<- data.frame(model = "log",augment(log_mod, newdata = newdata))
# log_pred$.fitted <- exp(log_pred$.fitted) 

nl_pred<- data.frame(model = "nonlinear",augment(nl_mod, newdata = newdata))
nl_pred$.se.fit<-NA

all.pred<- rbind(lm_pred,poly_pred,log_pred,nl_pred)

ggplot(data = FishAge) +
  geom_point(aes(x = Age, y = FishLength), size = 1, alpha = 0.35, colour = "red") +
  geom_line(data=all.pred, aes(x = Age, y = .fitted, colour = model)) +
  theme_bw()
```

Lets look at the distribution of the residuals of each model to the actual data.  What would you expect to see?


```{r}
resid_lm<-data.frame(model = "linear",resid = augment(lm_mod)$.resid)
resid_log<-data.frame(model = "log",resid = augment(log_mod)$.resid)
resid_poly<-data.frame(model = "poly",resid = augment(poly_mod)$.resid)
resid_nl<-data.frame(model = "nonlinear",resid = augment(nl_mod)$.resid)


all_resid<-rbind(resid_lm, resid_log, resid_poly, resid_nl)
head(all_resid)
glimpse(all_resid)

ggplot(data = all_resid) +
  geom_violin(aes(x = model, y  =resid, fill = model)) + 
  theme_bw()
```

## Experiments, Events, and Probability

- In probability theory, we are concerned with the occurence of events that can be thought of as outcomes of experiments

- The probability of event A occurring is \\( Pr \{ A \} \\) = probability that the event occurs 

 - The _Frequentist_ interpretation of probability \\( Pr \{ A \} \\) is the proportion of A outcomes as the total number of trials in an experiment goes to infinity. 
 
 Coin flipping example:
 
 For example, it can be demonstrated that the proportion of heads from a series of fair coin  flips will approach the constant 0.5 as the number of trials grows large, that is, \\( Pr \{ Head \} \\) = 0.5
 
```{r}

## Binomial distribution
?rbinom
set.seed(12345)

N_flips<-data.frame(N = c(1:5000))
N_flips$N_heads <- unlist(lapply(lapply(N_flips$N, rbinom, size = 1, prob = 0.50), sum))
N_flips$prop <- N_flips$N_heads/N_flips$N

head(N_flips)

```

If we look at the first 100, you can see 

```{r}
ggplot(data = N_flips) +
  geom_line(aes(x = N, y = prop)) + 
  geom_hline(aes(yintercept = 0.5), colour = "red", linetype = "dotted") +
  coord_cartesian(xlim = c(0, 100)) +
  theme_bw()

```

Lets look at the whole range

```{r}
ggplot(data = N_flips) +
  geom_line(aes(x = N, y = prop)) + 
  geom_hline(aes(yintercept = 0.5), colour = "red", linetype = "dotted") +
  coord_cartesian(xlim = c(0, 5000)) +
  theme_bw()
```

- The _Bayesian_ interpretation of probability is the degrees of belief. For a Bayesian, \\( Pr\{A\}\\) is a measure of certainty; a quantication of an investigator’s belief that A is true.
 
 - Differences in Frequentist and Bayesian perspectives are most important pertaining to inferential procedures (e.g., parameter estimation and hypothesis testing). They are irrelevant to the mathematical principles of probability.