---
title: "Applied Multivarite:  Intro to R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome back to another semester of our UseR.  

[R script from class](https://github.com/chrischizinski/SNR_R_Group/blob/master/R/2017-08-25-Intro_dataimport.R)

[Powerpoint from class](/SNR_R_Group/figs/Intro to programming.pdf)


## Weekly challenge

### The data

Work with the ecology data set from datacarpentry.  An explanation of the dataset can be found [here](http://www.datacarpentry.org/R-ecology-lesson/02-starting-with-data.html)

```{r message = FALSE, warning = FALSE}
library(tidyverse)

mydata <- read_csv("https://ndownloader.figshare.com/files/2292169")

glimpse(mydata)
```

### In-class challenges

**1. Find the average hindfoot_length**

```{r}
mean(mydata$hindfoot_length)

# Notice the answer is NA.  This is because NAs are in the data and thus need to be removed. This can be accomplished in two ways

# 1. Remove the NAs from the data and create a new object and then take the mean

hindfoot_length.rev <- mydata$hindfoot_length[!is.na(mydata$hindfoot_length)]
avg_foot <- mean(hindfoot_length.rev)
avg_foot

# 2. Use  na.rm option within the mean function
?mean
avg_foot <-mean(mydata$hindfoot_length, na.rm = TRUE)
avg_foot
```

**2. How many are above and below average**

```{r}
#step 1 - find average
avg <- mean(mydata$hindfoot_length, na.rm=T)
#step 2- index to get only value in number below
lessthan <- mydata$hindfoot_length < avg
head(lessthan, 25) # display the first 25 elements

greaterthan <- mydata$hindfoot_length > avg

#step 3 - count the number of rows
sum(lessthan, na.rm=T)
sum(greaterthan, na.rm=T)

```

### Take-home challenges

**1. What are the names of the plot types (treatments) in the experiment?**

**2. How many species caught?**

**3. How many species of birds? Rodents?**

**4. Average weight of Male Rodents? **

**5. Average weight of Female Rodents?**

