---
title: "Applied Multivariate:  Identifying differences between groups, part 2"
output:
  html_notebook: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6)
```

```{r message = FALSE}
library(vegan) 
library(gridExtra)
library(tidyverse)
```


## MANOVA and MANCOVA

Two common multivariate regression models are the multivariate analysis of variance (MANOVA; an extension of the univariate ANOVA) when all predictor variables are categorical, and the multivariate analysis of covariance (MANCOVA; extension of univariate ANCOVA) when predictor variables are categorical and continuous. 

### Load the data

```{r}
bryc_data <- read.table("http://ecology.msu.montana.edu/labdsv/R/labs/lab1/bryceveg.R")


bryc_data %>% rownames_to_column("site") -> bryc_data

bryc_site_data <- read.table("http://ecology.msu.montana.edu/labdsv/R/labs/lab2/brycesite.R")

bryc_site_data %>% 
  rownames_to_column("site") %>% 
  full_join(bryc_data, by = "site") -> bryc_site_data

glimpse(bryc_site_data)


```

### Using MANOVA and MANCOVA

### MANOVA

-Independent variable is a categorical variable
- Interpret like ANOVA
- degfault test statistic is "Pillai" but "Wilks", "Hotelling-Lawley", "Roy" are also available. See `?summary.manova`. Wilks' statistic is most popular in the literature, but the default Pillai–Bartlett statistic is recommended by [Hand and Taylor (1987)](https://www.crcpress.com/Multivariate-Analysis-of-Variance-and-Repeated-Measures-A-Practical-Approach/Hand-Taylor/p/book/9780412258008)

As a **reminder**:

-R provides Type I sequential SS, not the default Type III marginal SS reported by SAS and SPSS. In a nonorthogonal design with more than one term on the right hand side of the equation order will matter (i.e., A+B and B+A will produce different results)!
  - Type I sequential SS:   tests the main effect of factor A, followed by the main effect of factor B after the main effect of A, followed by the interaction effect AB after the main effects.
  - Type II: tests for each main effect after the other main effect. No significant interaction is assumed (in other words, you should test for interaction first (SS(AB | A, B)) and only if AB is not significant, continue with the analysis for main effects).
- Type III:  tests for the presence of a main effect after the other main effect and interaction. This approach is therefore valid in the presence of significant interactions.  If the interactions are not significant, type II gives a more powerful test.

- NOTE: when data is balanced, the factors are orthogonal, and types I, II and III all give the same results.

- If you wish to find ways to produce Type II or III SS then you will need to find a package to help.  For example, `regr0` has a `drop1()` function for `mlm` models.  Install via `install.packages("regr0", repos="http://R-Forge.R-project.org")`

```{r}
man_out <- manova(cbind(junost,ameuta,arcpat,arttri,atrcan) ~ pos,data = bryc_site_data)
summary(man_out)

```

### MANCOVA

- When you have a categorical and continuous independent variables

```{r}
man_out <- manova(cbind(arcpat, artarb, carrss, purtri,sticom) ~ pos + elev, data = bryc_site_data)
summary(man_out)

```

## ANOSIM

Analysis of similarities (ANOSIM) provides a way to test statistically whether there is a significant difference between two or more groups of sampling units.

The R-statistic in ANOSIM is a ratio between within-group and between-group dissimilarities

```{r}
# create the y variable 
y<-cbind(bryc_site_data$arcpat,bryc_site_data$artarb,bryc_site_data$carrss,bryc_site_data$purtri,bryc_site_data$sticom)

# convert to an x variable
x <- as.factor(bryc_site_data$pos)

# analysis
anosim_out <- anosim(y, x, distance = "euclidean", permutations = 999)
summary(anosim_out)

perms<-permustats(anosim_out)
summary(perms)
densityplot(perms)
anosim_out$perm
```

R value is supposed to vary between 0 and 1 (not between -1 and +1) but you can obtained negative values but they are always close to 0. R value close to 1 indicates high separation between levels of your factor (e.g. control vs treatment samples), while R value close to 0 indicate no separation between levels of your factor. 