---
title: "Applied Multivariate:  Identifying differences between groups, part 2"
output:
  html_notebook: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6)
```

```{r message = FALSE}
library(vegan) 
library(gridExtra)
library(tidyverse)
```


## MANOVA and MANCOVA

Two common multivariate regression models are the multivariate analysis of variance (MANOVA; an extension of the univariate ANOVA) when all predictor variables are categorical, and the multivariate analysis of covariance (MANCOVA; extension of univariate ANCOVA) when predictor variables are categorical and continuous. 

### Load the data

```{r}
bryc_data2 <- read.table("http://ecology.msu.montana.edu/labdsv/R/labs/lab1/bryceveg.R")


bryc_data2 %>% rownames_to_column("site") -> bryc_data

bryc_site_data_og <- read.table("http://ecology.msu.montana.edu/labdsv/R/labs/lab2/brycesite.R")

bryc_site_data_og %>% 
  rownames_to_column("site") %>% 
  full_join(bryc_data, by = "site") -> bryc_site_data

head(bryc_site_data)


```

### Using MANOVA and MANCOVA

### MANOVA

-Independent variable is a categorical variable
- Interpret like ANOVA
- degfault test statistic is "Pillai" but "Wilks", "Hotelling-Lawley", "Roy" are also available. See `?summary.manova`. Wilks' statistic is most popular in the literature, but the default Pillai–Bartlett statistic is recommended by [Hand and Taylor (1987)](https://www.crcpress.com/Multivariate-Analysis-of-Variance-and-Repeated-Measures-A-Practical-Approach/Hand-Taylor/p/book/9780412258008)

As a **reminder**:

-R provides Type I sequential SS, not the default Type III marginal SS reported by SAS and SPSS. In a nonorthogonal design with more than one term on the right hand side of the equation order will matter (i.e., A+B and B+A will produce different results)!
  - Type I sequential SS:   tests the main effect of factor A, followed by the main effect of factor B after the main effect of A, followed by the interaction effect AB after the main effects.
  - Type II: tests for each main effect after the other main effect. No significant interaction is assumed (in other words, you should test for interaction first (SS(AB | A, B)) and only if AB is not significant, continue with the analysis for main effects).
- Type III:  tests for the presence of a main effect after the other main effect and interaction. This approach is therefore valid in the presence of significant interactions.  If the interactions are not significant, type II gives a more powerful test.

- NOTE: when data is balanced, the factors are orthogonal, and types I, II and III all give the same results.

- If you wish to find ways to produce Type II or III SS then you will need to find a package to help.  For example, `regr0` has a `drop1()` function for `mlm` models.  Install via `install.packages("regr0", repos="http://R-Forge.R-project.org")`

```{r}
man_out <- manova(cbind(junost,ameuta,arcpat,arttri,atrcan) ~ pos,data = bryc_site_data)
summary(man_out)

```

### MANCOVA

- When you have a categorical and continuous independent variables

```{r}
man_out <- manova(cbind(arcpat, artarb, carrss, purtri,sticom) ~ pos + elev, data = bryc_site_data)
summary(man_out)

```

## ANOSIM

Analysis of similarities (ANOSIM) provides a way to test statistically whether there is a significant difference between two or more groups of sampling units.

The R-statistic in ANOSIM is a ratio between within-group and between-group dissimilarities.  R is interpreted like a correlation coefficient and is a measure of ‘effect size’

R value is supposed to vary between 0 and 1 (not between -1 and +1) but you can obtained negative values but they are always close to 0. R value close to 1 indicates high separation between levels of your factor (e.g. control vs treatment samples), while R value close to 0 indicate no separation between levels of your factor. 

```{r}
# create the y variable 
y<-cbind(bryc_site_data$arcpat,bryc_site_data$artarb,bryc_site_data$carrss,bryc_site_data$purtri,bryc_site_data$sticom)

# convert to an x variable
x <- as.factor(bryc_site_data$pos)

# analysis
anosim_out <- anosim(y, x, distance = "euclidean", permutations = 999)
summary(anosim_out)

```
The Dissimilarity ranks between and within classes are the  ranks of over all dissimilarities for
quantiles of the dissimilarities in each "class". The Between row is for
dissimilarities for pairs of samples in different groups.  Other rows are the within group pairs.   


ANOSIM assumes Between and within (the other rows) should have similar distributions under the Null hypothesis.  The table is presented so the user can evaluate whether the assumptions are being met.

P-value is calculated as the probability of an R larger through Monte carlo permutations.

```{r}
perms<-permustats(anosim_out) # collect the permutation values

perm_data<- data.frame(perm_value = perms$permutations) # convert to a dataframe

perm_data$greater <- ifelse(perm_data$perm_value>=perms$statistic ,1,0)

sum(perm_data$greater)/length(perm_data$greater) # calculate p value

# Illustrate with a plot

ggplot(data = perm_data) +
  geom_density(aes(x = perm_value)) +
  geom_point(aes(x = perm_value, y = 0.5 ), position = "jitter", alpha = 0.25, color = "red") +
  geom_vline(aes(xintercept = perms$statistic ), linetype = "dashed") +
  annotate("text", x = perms$statistic, y = 15, label = paste("R =", round(perms$statistic, digits = 3)), hjust = 0) + 
  coord_cartesian(ylim = c(0,20), expand = F) +
  theme_classic()


?mrpp


```

## Multiresponse permutation procedure (MRPP)

MRPP provides a test of whether there is a significant difference between two or more groups of sampling units. Difference may be one of location (differences in mean) or one of spread (differences in within-group distance).

MRPP provides little more than a test statistic of effect size and p value.  

*A* is proportion of the distances explained by group identity; a value analogous to a coefficient of determination in a linear model.
- A = 1 when all items are identical within groups
- A = 0 when within-group heterogeneity equals expectation by chance
- A < 0.1 common in ecology
- A > 0.3 is fairly high in ecology


```{r}
mrpp_out <- mrpp(y, x, distance = "euclidean", permutations = 999)
mrpp_out

mrpp_out$A  # measure of effect size
mrpp_out$Pvalue # p value
```

## Permutational Multivariate Analysis of Variance Using Distance Matrices (Adonis)

Analysis of variance using distance matrices — for partitioning distance matrices among sources of variation and fitting linear models (e.g., factors, polynomial regression) to distance matrices; uses a permutation test with pseudo-F ratios.

`adonis` and `adonis2` are robust methodologies

`adonis` performs a sequential test of terms. 
`adonis2` performs sequential, marginal and overall tests. 

```{r}
data(dune)
data(dune.env)

adon1_out<-adonis(dune ~ Management*A1, data = dune.env)
adon1_out

adon2_out <-adonis2(dune ~ Management*A1, data = dune.env, by = "terms", perm=999) # significance for each term (sequentially from first to last),
adon2_out

adon3_out <-adonis2(dune ~ Management*A1, data = dune.env, by = "margin", perm=999) #  marginal effects of the terms (each marginal term analysed in a model with all other variables)
adon3_out

adon4_out <-adonis2(dune ~ Management*A1, data = dune.env, by = NULL, perm=999) #   overall significance of all terms together.
adon4_out
```

```{r}

dune_bray<- vegdist(dune, "bray")
mod <-betadisper(dune_bray,dune.env$Management, type = "centroid")
mod

# extract the centroids and the site points in multivariate space.  
centroids<-data.frame(grps=rownames(mod$centroids),data.frame(mod$centroids))
vectors<-data.frame(grps=mod$group,data.frame(mod$vectors))

names(vectors)<-paste("v.",names(vectors),sep="")

# to create the lines from the centroids to each point we will put it in a format that ggplot can handle
vectors %>% 
  left_join(centroids, by = c("v.grps" = "grps")) %>% 
  rename(grps = v.grps) %>% 
  select (grps, PCoA1, PCoA2, v.PCoA1, v.PCoA2)-> seg.data

all_data <- data.frame(grps = "all",
           PCoA1 = mean(seg.data$v.PCoA1),
           PCoA2 = mean(seg.data$v.PCoA2),
           v.PCoA1 = seg.data$v.PCoA1,
           v.PCoA2 = seg.data$v.PCoA2)

seg.data<-rbind(seg.data, all_data)

ggplot(data = seg.data) + 
  geom_segment(data=seg.data,aes(x=PCoA1,xend=v.PCoA1,y=PCoA2,yend=v.PCoA2),alpha=0.30) +
  geom_point(aes(x = PCoA1, y = PCoA2), colour = "red") +
  geom_point(aes(x = v.PCoA1,y=v.PCoA2)) +
  facet_wrap(~grps, nrow = 1)

```

