---
title: "Intro to Multivariate Stats"
output:
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Difference between Transformations and Standardizations

- Transformations are applied to each element in a matrix
- Standardization adjust elements in a matrix by a row or column statistic 

### Create some data

```{r}

rawdata <- matrix(c(1,1,1,3,3,1,
                    2,2,4,6,6,0,
                    10,10,20,30,30,0,
                    3,3,2,1,1,0,
                    0,0,0,20,0,0), ncol = 6, byrow = TRUE)
colnames(rawdata) <- paste("species",toupper(letters[1:6]), sep = "_")

rawdata
```

### Calculating row and column statistics

#### Rows

```{r}
# Row sums
rowSums(rawdata)

apply(rawdata, 1, sum)

# Max values
apply(rawdata, 1, max)
```

#### Columns 

```{r}
# Sums
apply(rawdata, 2, sum)
colSums(rawdata)

# Max
apply(rawdata, 2, max)
```

### Monotonic transformations 

#### Log transformations 

- Useful for when you have a wide spread in data values

- Ir is important that you add 1 to values to account for zeros `log10(x+1)`

```{r}
logdata <- apply(rawdata , c(1,2), function(x) log10(x + 1))
```

```{r message = FALSE}
library(tidyverse)

hemlock <- read_csv("https://raw.githubusercontent.com/chrischizinski/SNR_R_Group/master/data/hemlock_cover.csv")



hemlock$logTsuga<- log10(hemlock$Tsuga.canadensis +1) # log transform 

glimpse(hemlock)

```

```{r}
ggplot(data = hemlock) + 
  geom_histogram(aes(Tsuga.canadensis), binwidth = 5, colour = "black", fill = "dodgerblue") + 
  coord_cartesian(ylim = c(0, 30), expand = FALSE) +
  theme_bw()
```

```{r}
ggplot(data = hemlock) + 
  geom_histogram(aes(logTsuga), bins = 30, colour = "black", fill = "red") + 
  coord_cartesian(ylim = c(0, 30), expand = FALSE) +
  theme_bw()
```

#### Power tranformations

- Square root transformation is most often used for Poisson type date (count data)
- Greater the power, the greater the compression of the data
- Flexible for a wide range of data
- Applied when the data is > 0

##### Write power function

```{r}
pwr_trans <- function(x, trans){
  x <- ifelse(x>0,x^(1/trans),0)
  return(x)
}

pwr_trans(25,2)

pwr_trans(0,2)

```

#### Display the effect of the power function

```{r}
newdata <- data.frame(x = 0:100, 
                      cubic = pwr_trans(x=0:100, trans = 3),
                      power10 = pwr_trans(x=0:100, trans = 10))

head(newdata)
```

```{r}
ggplot(data = newdata) +
  geom_line(aes(x = x, y = cubic), size = 1, colour = "blue") +
  geom_line(aes(x = x, y = power10), size = 1, colour = "red") +
  labs(y = "Value") +
  coord_cartesian(xlim = c(0,100.5), ylim = c(0,5), expand = F)+
  theme_classic()
  
```


#### Presence absence transformation

- Transforms quantitative data to non-quanitative (binary)
- Applicable to species data
- Most useful when there is not a lot of quantitative data available (i.e., LOTS of zeros)
- Severe transformation (i.e., loose lots of information) 

```{r}
library(vegan)
decostand(rawdata, method = "pa")
```

#### Arcsine transformation

Please NOTE: [The arcsine is asinine: the analysis of proportions in ecology](http://onlinelibrary.wiley.com/doi/10.1890/10-0340.1/abstract)

- Transformations on proportion data (0-1)
- Useful when you have a positive skew in data
    - Spreads the end of the scale while compressing the middle 

### Standardizations

### Sums 
- Can be applied to any range of x
- Output will range 0 - 1
- Converts values to a relative value (equalizes the area under the curve) 
- Useful when you have large difference in total abundance

#### Rows

```{r}
ttl_species <- apply(rawdata, 1, sum)
rowprop_data <- rawdata / ttl_species
  
rowprop_data

decostand(rawdata, margin = 1, method = "total")
```

#### Columns

```{r}
colprop_data <- rawdata %*% diag(1/apply(rawdata,2,sum))
```

## Take home challenge 

1) Input the data that Mike W. showed `mydata <- read_csv("https://ndownloader.figshare.com/files/2292169")`
2) Calculate the number of unique individuals per species per site per year per month for Rodents
3) Calculate the maximum number of unique individuals per species per site per year for Rodents
4) Calculate the max individuals across years per species per site for Rodents
5) Create a site by species matrix composed of those max values
6) Calculate by row total standardization
7) Plot a bar chart of each site and relative number of a few (3) species

Send to cchizinski2@unl.edu