---
title: "Applied Multivariate:  Ordination"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggrepel)
library(vegan)
library(tidyverse)
```

## Ordination

Ordination is the process of placing multivariate data in ordination space such that similar objects are near each other and dissimilar objects are farther from each other (think of this like a map)


### Principal Component Analysis (PCA)

-  PCA is used for exploratory data analysis (Ramette 2007) or developing composite values for more complex methods 
- reduces data into so-called components through linear combinations of variables that maximize the variance explained in the original data
  - Using Euclidean distances, each successive composite principal component is orthogonal to the last (reducing correlation) and explains less variation than the previous component (i.e. the first component explains the most variation).  

### Data

PCA requires:
- single set of interdependent variables 
- continuous, categorical, or count variables (preferably all continuous)
- no missing data 
- ideally there should be more samples (rows) than number of variables (columns) 

```{r message = FALSE}
## Data is from a sample dataset is taken from research conducted in the Drift Creek Basin in the central Oregon Coast Range during 1990 (McGarigal and McComb, unpubl. data).  Downloaded from http://www.umass.edu/landeco/teaching/multivariate/labs/multivariate_labs.html

bird_hab <- read_csv("https://raw.githubusercontent.com/chrischizinski/SNR_R_Group/master/data/birdhab.csv")
glimpse(bird_hab)

# limit the data to just the birds and just the habitat

bird_hab %>% select(AMGO:WWPE) -> bird_rip
bird_hab %>% select(VAREA:CONEDGE) -> hab_rip

# remove birds that have frequency less than five
decostand(bird_rip, "pa") %>% 
  gather(species, pres, AMGO:WWPE) %>% 
  group_by(species) %>% 
  summarise(TTL = sum(pres)) %>% 
  filter(TTL < 5) -> infreq_birds

bird_rip %>% 
  select(-one_of(infreq_birds$species)) -> bird_rip.red
  
```

### Conduct the PCA

We will use the `prcomp()` function

PCA is sensitive to the scale of the data 

If all the data are not measured on the same scale,using covariance means that the result will be determined mostly by the variable with the largest values, as it will have the highest variance. 

Using a correlation matrix treats all variables the same (standardized to mean=0 and std. dev=1).

```{r}
rip_bird.pca<-prcomp(bird_rip.red,scale=TRUE)
names(summary(rip_bird.pca))
```

#### Scree

```{r}

# extract the variances
data_scree <- data.frame(comp = 1:49, var= summary(rip_bird.pca)$sdev^2)

# plot them out
ggplot(data = data_scree) + 
  geom_line(aes(x = comp, y = var)) +
  geom_point(aes(x = comp, y = var)) + 
  theme_classic()
           
```

#### Plotting as a biplot

The points on the biplot represent observations positioned in ordination space by their principal components such that component 1 is represented on the x-axis and component 2 on the y–axis 

interpretation of ordination is dependent on whether a distance biplot (intersample relationships; scaling 1) or a correlation biplot (interspecies correlations; scaling 2) is used, because observation scores are rescaled as a function of the scaling choice

```{r}
scale = 1  # covariance biplot (scale = 1), form biplot (scale = 0)
var.scale = scale #scale factor to apply to variables
obs.scale = 1 - scale #scale factor to apply to observations
circle.prob = 0.69

nobs.factor <- sqrt(nrow(rip_bird.pca$x) - 1) 
d <- rip_bird.pca$sdev
u <- sweep(rip_bird.pca$x, 2, 1 / (d * nobs.factor), FUN = '*')
v <- rip_bird.pca$rotation

# Scores
df.u <- as.data.frame(sweep(u, 2, d^obs.scale, FUN='*'))

  # Directions
  v <- sweep(v, 2, d^var.scale, FUN='*')
  df.v <- as.data.frame(v)

  df.u <- df.u * nobs.factor

  # Scale the radius of the correlation circle so that it corresponds to 
  # a data ellipse for the standardized PC scores
  r <- sqrt(qchisq(circle.prob, df = 2)) * prod(colMeans(df.u^2))^(1/4)

  # Scale directions
  v.scale <- rowSums(v^2)
  df.v <- r * df.v / sqrt(max(v.scale))
  df.v %>% rownames_to_column("species") -> df.v

soi <- c("BEKI","HAFL","AMRO","WIFL")

ggplot(data = df.u) +
  geom_point(aes(x = PC1, y = PC2)) + 
  geom_text(aes(x = PC1, y = PC2, label = paste("site",rownames(df.u))), alpha = 0.25, hjust = 0.5) +
  geom_segment(data = df.v %>% filter(species %in% soi), aes(x = 0, y = 0, xend = PC1, yend = PC2), 
               arrow = arrow(length = unit(1/2, 'picas')), 
               color = muted('red')) + 
  geom_label_repel(data = df.v %>% filter(species %in% soi), aes(x = PC1, y = PC2, label = species)) +
  theme_classic() 

```

## Principal Coordinate Analysis (PCoA)

Principal Coordinate Analysis (PCoA), or metric multidimensional scaling, is similar to a Principal Component Analysis but can accept distance measures other than Euclidean.  

PCoA using Euclidean distances produces the same results as a PCA

```{r}
dim(bird_rip.red)

bray_bird <- vegdist(bird_rip.red, "bray")

rip_bird.pcoa<-cmdscale(bray_bird, k = 2)
rip_bird.pcoa
```

```{r}
scores(rip_bird.pcoa) %>% as.data.frame() ->pcoa_data
ggplot(data = pcoa_data) +
  geom_point(aes(x = Dim1, y = Dim2)) + 
  geom_text(aes(x = Dim1, y = Dim2, label = rownames(pcoa_data)), alpha = 0.25, hjust = 0.5) +
  theme_classic() 
```

# Correspondance Analysis

most frequently use CA to compare the similarity (correspondence) between sites (samples) and species abundances to represent these similarities in ordination space 

Correspondence analysis uses chi-square distances, which requires observations be positive and measured on the same scale 

CA is a much better and more robust method for community ordination than principal components analysis. 

However, with long ecological gradients it suffers from some drawbacks or “faults” which were corrected in Detrended Correspondence Analysis (DCA):
1. Single long gradients appear as curves or arcs in ordination: the solution is to detrend the later axes by making their means equal along segments of previous axes.
2. Sites are packed more closely at gradient extremes than at the center: the solution is to rescale the axes to equal variances of species scores.
3. Rare species seem to have an unduly high influence on the results: the solution is to downweight rare species.

```{r}
bird_cca<-cca(bird_rip.red)
bird_cca

bird_cca$CA$eig/sum(bird_cca$CA$eig)

bird_cca$CA$u #site eigenvectors (‘u’)
bird_cca$CA$v #species eigenvectors (‘v’)

summary(bird_cca, scaling = 1, axes = 3)


# devtools::install_github("gavinsimpson/ggvegan")
# library(ggvegan)



```

CA has several conventions for scaling the sample and species scores. Recall that CA is a weighted averaging method (at least conceptually). In scaling type 1, the locations of the sample plots are calculated weighted averages of species scores; i.e., as the means of the species scores for species that occur in the plot, weighted by species abundance. Thus,
species scores often lie outside the range of plot scores in the graph. In scaling type 2, species scores
are calculated as weighted averages of samples scores; i.e., the means of the plot scores in which the
species occurs, weighted by its abundance in each plot.

```{r}
score_1<-scores(bird_cca, scaling = 1)
species_1 <-score_1$species %>% as.data.frame() %>% rownames_to_column("species")
sites_1 <-score_1$sites %>% as.data.frame() %>% rownames_to_column("site")

ggplot() +
  geom_point(data = species_1,aes(x = CA1, y = CA2), shape = 16, colour = "red") + 
  geom_text(data = species_1,aes(x = CA1, y = CA2, label = species), alpha = 0.25, hjust = 1) +
  geom_point(data = sites_1,aes(x = CA1, y = CA2), shape = 24, colour = "blue") + 
  theme_classic()

score_2<-scores(bird_cca, scaling = 2)
species_2 <-score_1$species %>% as.data.frame() %>% rownames_to_column("species")
sites_2 <-score_1$sites %>% as.data.frame() %>% rownames_to_column("site")

ggplot() +
  geom_point(data = species_2,aes(x = CA1, y = CA2), shape = 16, colour = "red") + 
  geom_text(data = species_2,aes(x = CA1, y = CA2, label = species), alpha = 0.25, hjust = 1) +
  geom_point(data = sites_2,aes(x = CA1, y = CA2), shape = 24, colour = "blue") + 
  theme_classic()
```

this curve may be a mathematical consequence that both maximizes the difference between species and minimizes correlation between axes 

```{r}
bird_dca<-decorana(bird_rip.red)
names(bird_dca)

species_1 <-bird_dca$cproj %>% as.data.frame() %>% rownames_to_column("species")
sites_1 <-bird_dca$rproj %>% as.data.frame() %>% rownames_to_column("site")

ggplot() +
  geom_point(data = species_1,aes(x = DCA1, y = DCA2), shape = 16, colour = "red") + 
  geom_text(data = species_1,aes(x = DCA1, y = DCA2, label = species), alpha = 0.25, hjust = 1) +
  geom_point(data = sites_1,aes(x = DCA1, y = DCA2), shape = 24, colour = "blue") + 
  theme_classic()
```

## Nonmetric multidimensional scaling

Non-metric multidimensional scaling has the best performance characteristics of the unconstrained ordination methods in ecology 

While most ordination methods attempt to maintain the distance measures between points in multivariate space (i.e., the distance between points in an ordination represent the distance measures between points), there are situations where the exact distance is immaterial and the rank order of the relationships is sufficient

NMDS is able to use a variety of distance measures.  

Nonmetric multidimensional scaling, as opposed to other ordination methods, does not rely on eigenanalysis and does not maximize the variation explained by each axis. 

NMDS axes are arbitrary and can be flipped or rotated.

Unlike the eigenanalysis methods described in PCA, CA, and PCoA, the number of axes are described a priori in NMDS.  

The points are then oriented around the a priori specified number of axes at random locations.

The distances from the random locations are compared to the distances in the original data using isotonic regression with a stress function (ranges between 0 and 1).

The points are iteratively moved in the direction of decreasing stress until the lowest stress value is found.  

```{r}

max(bird_rip.red)
min(bird_rip.red)

bird_trans<-decostand(bird_rip.red, "total")

bird_nmds <- metaMDS(bird_trans, k = 3, distance = "euclidean", autotransform = FALSE, try = 50)

```

```{r}
species_scores <- scores(bird_nmds, "species") %>%  as.data.frame() %>% rownames_to_column("species")
site_scores <- scores(bird_nmds, "site") %>%  as.data.frame() %>% rownames_to_column("site")

ggplot() +
  geom_point(data = species_scores,aes(x = NMDS1, y = NMDS2), shape = 16, colour = "red") + 
  geom_text(data = species_scores,aes(x = NMDS1, y = NMDS2, label = species), alpha = 0.25, hjust = 1) +
  geom_point(data = site_scores,aes(x = NMDS1, y = NMDS2), shape = 24, colour = "blue") + 
  theme_classic()

```


### Fitting environmental variables

```{r}
head(hab_rip)

edge.sf <- ordisurf(bird_nmds ~ hab_rip$TEDGE, plot = TRUE)

extract.xyz <- function(obj) {
    xy <- expand.grid(x = obj$grid$x, y = obj$grid$y)
    xyz <- cbind(xy, c(obj$grid$z))
    names(xyz) <- c("x", "y", "z")
    return(xyz)
}


contour.vals <- extract.xyz(obj = edge.sf)
head(contour.vals)

ggplot(data = contour.vals) + 
   geom_point(data = species_scores,aes(x = NMDS1, y = NMDS2), shape = 16, colour = "red") + 
  geom_text(data = species_scores,aes(x = NMDS1, y = NMDS2, label = species), alpha = 0.25, hjust = 1) +
  geom_point(data = site_scores,aes(x = NMDS1, y = NMDS2), shape = 24, colour = "blue", fill = "blue", alpha = 0.5) + 
  geom_contour(aes(x = x, y = y, z = z))  + 
  theme_classic()



```

