---
title: "Identifying differences between groups"
output:
  html_notebook: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6)
```

```{r message = FALSE}
library(vegan) 
library(gridExtra)
library(tidyverse)
```


## Linear discriminant analysis

- Often we have groups that we have defined *a priori* before doing an analysis and we seek to understand what makes those groups different

- Linear discrimination analysis (LDA), similar to multinomial logistic regression, attempts to find linear combinations of variables that best separate groups when predicting two or more dependent variables, using continuous independent variables. 

### Load the data

We will use the data set `morph_data.csv` on [github](). The data consisted of  8 measurements in all 5
larval stages and the adults of the 6 species of the waterstrider 
genus Limnoporus (Insecta: Heteroptera: Gerridae). Data was originally sourced from [morphometry datasets](http://life.bio.sunysb.edu/morph/data/datasets.html).

The variables are:  Species, stage (1-5 for larvae, 6 for adults), sex (m male, f female, u undetermined), antseg1 - antseg4 (lengths of 1st to 4th antennal segments), midfem, midtib, hindfem, hindtib (lenghts of middle and hind femora and tibiae).
The values are raw measurements in millimeters.

```{r}
morph_data <-read_csv("https://raw.githubusercontent.com/chrischizinski/SNR_R_Group/master/data/morph_data.csv")

glimpse(morph_data)

```

### Manipulate the data

For the purpose of this excercise lets limit the data to one species and only the measurements of the antennae. 

```{r}
morph_data %>% 
  select(Species:antseg4, -sex) %>% 
  filter(Species == "canali") %>% 
  mutate_at(vars(antseg1:antseg4), scale)-> canal_data

head(canal_data)
```

### Conduct the linear discriminant analysis using `MASS::lda()`

```{r}
canal_lda<-MASS::lda(stage~antseg1 + antseg2 + antseg3 + antseg4, data = canal_data)

canal_lda
names(canal_lda)
```

### Explanation of the output

#### Prior

From the raw data, the break down in proportions of the number of individuals in each group

```{r}
canal_lda$prior
```

#### Group means
The mean value for each of the variables for each group

```{r}
canal_lda$means
```

#### Coefficients

These are the coffeicients to describe the linear discrimination. For example: LD1 = `r canal_lda$scaling[1]` x Group1 + `r canal_lda$scaling[2]` x Group2 + `r canal_lda$scaling[3]` x Group3 + ...

```{r}
canal_lda$scaling

```

You can also standardize the coefficients (loadings), which provides the relative strength of each factor on linear discriminant analysis

```{r}
loadings <-canal_lda$scaling
col.ss <- colSums(loadings^2)
sweep(loadings,2,sqrt(col.ss),"/")
```


### Variation expained

The amount of variation explained by the each of the linear discriminant axis

```{r}
prop = canal_lda$svd^2/sum(canal_lda$svd^2)
prop
```


### Graphing the results

```{r}

# extracting the data
lda_pred <- predict(canal_lda)
lda_scores<-data.frame(stage = lda_pred$class, lda_pred$x)

#developing hulls for each stage group

hull1<- lda_scores[lda_scores$stage ==1,][chull(lda_scores[lda_scores$stage ==1,c("LD1", "LD2")]),]
hull2<- lda_scores[lda_scores$stage ==2,][chull(lda_scores[lda_scores$stage ==2,c("LD1", "LD2")]),]
hull3<- lda_scores[lda_scores$stage ==3,][chull(lda_scores[lda_scores$stage ==3,c("LD1", "LD2")]),]
hull4<- lda_scores[lda_scores$stage ==4,][chull(lda_scores[lda_scores$stage ==4,c("LD1", "LD2")]),]
hull5<- lda_scores[lda_scores$stage ==5,][chull(lda_scores[lda_scores$stage ==5,c("LD1", "LD2")]),]
hull6<- lda_scores[lda_scores$stage ==6,][chull(lda_scores[lda_scores$stage ==6,c("LD1", "LD2")]),]

all_hulls <- rbind(hull1, hull2, hull3, hull4, hull5, hull6)

#plot using gglot2

ggplot(data = lda_scores) + 
  geom_point(aes(x = LD1, y = LD2, color = stage), size = 2) +
  geom_polygon(data = all_hulls,aes(x = LD1, y = LD2, color = stage, fill = stage), alpha = 0.25) +
  coord_equal() + 
  theme_classic()
```

### Assessing the accuracy of the predicitions

```{r}
ct <- table(canal_data$stage, lda_pred$class)
ct
# percent correct for each stage
diag(prop.table(ct, 1))

# total percent correct
sum(diag(prop.table(ct)))
```

## MANOVA and MANCOVA

Two common multivariate regression models are the multivariate analysis of variance (MANOVA; an extension of the univariate ANOVA) when all predictor variables are categorical, and the multivariate analysis of covariance (MANCOVA; extension of univariate ANCOVA) when predictor variables are categorical and continuous. 

### Load the data

```{r}
bryc_data <- read.table("http://ecology.msu.montana.edu/labdsv/R/labs/lab1/bryceveg.R")


bryc_data %>% rownames_to_column("site") -> bryc_data

bryc_site_data <- read.table("http://ecology.msu.montana.edu/labdsv/R/labs/lab2/brycesite.R")

bryc_site_data %>% 
  rownames_to_column("site") %>% 
  full_join(bryc_data, by = "site") -> bryc_site_data

glimpse(bryc_site_data)


```

### Using MANOVA and MANCOVA

### MANOVA

-Independent variable is a categorical variable
- Interpret like ANOVA
- degfault test statistic is "Pillai" but "Wilks", "Hotelling-Lawley", "Roy" are also available. See `?summary.manova`. Wilks' statistic is most popular in the literature, but the default Pillai–Bartlett statistic is recommended by [Hand and Taylor (1987)](https://www.crcpress.com/Multivariate-Analysis-of-Variance-and-Repeated-Measures-A-Practical-Approach/Hand-Taylor/p/book/9780412258008)

As a **reminder**:

-R provides Type I sequential SS, not the default Type III marginal SS reported by SAS and SPSS. In a nonorthogonal design with more than one term on the right hand side of the equation order will matter (i.e., A+B and B+A will produce different results)!
  - Type I sequential SS:   tests the main effect of factor A, followed by the main effect of factor B after the main effect of A, followed by the interaction effect AB after the main effects.
  - Type II: tests for each main effect after the other main effect. No significant interaction is assumed (in other words, you should test for interaction first (SS(AB | A, B)) and only if AB is not significant, continue with the analysis for main effects).
- Type III:  tests for the presence of a main effect after the other main effect and interaction. This approach is therefore valid in the presence of significant interactions.  If the interactions are not significant, type II gives a more powerful test.

- NOTE: when data is balanced, the factors are orthogonal, and types I, II and III all give the same results.

- If you wish to find ways to produce Type II or III SS then you will need to find a package to help.  For example, `regr0` has a `drop1()` function for `mlm` models.  Install via `install.packages("regr0", repos="http://R-Forge.R-project.org")`

```{r}
man_out <- manova(cbind(junost,ameuta,arcpat,arttri,atrcan) ~ pos,data = bryc_site_data)
summary(man_out)

```

### MANCOVA

- When you have a categorical and continuous independent variables

```{r}
man_out <- manova(cbind(arcpat, artarb, carrss, purtri,sticom) ~ pos + elev, data = bryc_site_data)
summary(man_out)

```

## ANOSIM

Analysis of similarities (ANOSIM) provides a way to test statistically whether there is a significant difference between two or more groups of sampling units.

The R-statistic in ANOSIM is a ratio between within-group and between-group dissimilarities

```{r}
# create the y variable 
y<-cbind(bryc_site_data$arcpat,bryc_site_data$artarb,bryc_site_data$carrss,bryc_site_data$purtri,bryc_site_data$sticom)

# convert to an x variable
x <- as.factor(bryc_site_data$pos)

# analysis
anosim_out <- anosim(y, x, distance = "euclidean", permutations = 999)
summary(anosim_out)

perms<-permustats(anosim_out)
summary(perms)
densityplot(perms)
anosim_out$perm
```

R value is supposed to vary between 0 and 1 (not between -1 and +1) but you can obtained negative values but they are always close to 0. R value close to 1 indicates high separation between levels of your factor (e.g. control vs treatment samples), while R value close to 0 indicate no separation between levels of your factor. 